The unphysicality of Hilbert spaces

### 2023-07-25 Articles/2023-HilbertQuantum/HilbertQuantum.tex
Fixed small typos in the proof and added two minor TODOs

\begin{proof}
Let $\psi, \phi \mathcal{H}$ be two normalized vectors. They will identify a two-dimensional subspace which can be thought, without loss of generality, as a qubit and therefore can be represented by a Bloch sphere. The below picture represents the intersection of the Bloch sphere with the plane identified by $\psi$ and $\phi$. Let $\rho = \frac{1}{2}|\psi_1\>\<\psi_1| + \frac{1}{2}|\psi_2\>\<\psi_2|$ be the density operator that represent an equal mixture of the two states. It will be represented by the midpoint of the chord connecting $\psi$ and $\phi$. Now take the line that goes through $\rho$ and the center of the sphere: $\rho$ can also be expressed as the mixture of the states $+$ and $-$ which, since they represent equal and opposite directions, form a basis. To diagonalize $\rho$, then, means to express it in terms of $+$ and $-$.

@@

If $\theta_{\psi\phi}$ is the angle between $\psi$ and $\phi$ on the Bloch sphere, we have
\begin{equation}
	|\langle \psi | \phi \rangle |^2 = \cos^2 \frac{\theta_{\psi\phi}}{2}.
\end{equation}
The angle is divided by two because the angle on the Bloch sphere (i.e. in physical space) is double the angle in the Hilbert space. For example, the angle between $z^+$ and $z^-$ on the Bloch sphere is $\pi$, but the inner product between them is zero (i.e. opposite directions in physical space correspond to orthogonal states) and therefore the angle in the Hilbert space is $\pi/2$.

@@

As $\rho$ is in a diagonal form, the entropy is given by
\begin{equation}\label{entropy}
	H(\rho) = H\left(\frac{1+|\langle\psi|\phi\rangle|}{2}, \frac{1-|\langle\psi|\phi\rangle|}{2}\right).
\end{equation}
If we define $p(\psi, \phi) = |\langle\psi|\phi\rangle|$ and $I(\psi, \phi) = H( \frac{1}{2}|\psi_1\>\<\psi_1| + \frac{1}{2}|\psi_2\>\<\psi_2|)$ we have
\begin{align}\label{entropy}
	f(p) & = - \frac{1+p}{2} \log \frac{1+p}{2} 
	- \frac{1-p}{2} \log \frac{1-p}{2} \\
	I(\psi, \phi) &= f(p(\psi, \phi)).
\end{align}
Note that $f$ restricted to the domain $[0,1]$, the domain on which probability is defined, is invertible.
\end{proof}

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact}\footnote{Maybe Prop. 3.115 in \cite{moretti_spectral_2017}?} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of functions with compact support on phase space gives us finite expectation value for all functions of positions and momentum.}

>>>>

\begin{proof}
Let $\psi, \phi \in \mathcal{H}$ be two normalized vectors. They will identify a two-dimensional subspace which can be thought, without loss of generality, as a qubit and therefore can be represented by a Bloch sphere. The below picture represents the intersection of the Bloch sphere with the plane identified by $\psi$ and $\phi$. Let $\rho = \frac{1}{2}|\psi\>\<\psi| + \frac{1}{2}|\phi\>\<\phi|$ be the density operator representing an equal mixture of the two states. It will be represented by the midpoint of the chord connecting $\psi$ and $\phi$. Now take the line that goes through $\rho$ and the center of the sphere: $\rho$ can also be expressed as the mixture of the states $+$ and $-$ which, since they represent equal and opposite directions, form a basis. To diagonalize $\rho$, then, means to express it in terms of $+$ and $-$.

@@

If $\theta_{\psi\phi}$ is the angle between $\psi$ and $\phi$ on the Bloch sphere, we have
\begin{equation}
	|\langle \psi | \phi \rangle |^2 = \cos^2 \frac{\theta_{\psi\phi}}{2}.
\end{equation}
The angle is divided by two because the angle on the Bloch sphere (i.e. in physical space) is double the angle in the Hilbert space. For example, the angle between $+$ and $-$ on the Bloch sphere is $\pi$, but the inner product between them is zero (i.e. opposite directions in physical space correspond to orthogonal states) and therefore the angle in the Hilbert space is $\pi/2$.

@@

As $\rho$ is in a diagonal form, the entropy is given by
\begin{equation}\label{entropy}
	H(\rho) = H\left(\frac{1+|\langle\psi|\phi\rangle|}{2}, \frac{1-|\langle\psi|\phi\rangle|}{2}\right).
\end{equation}
If we define $p(\psi, \phi) = |\langle\psi|\phi\rangle|$ and $I(\psi, \phi) = H( \frac{1}{2}|\psi\>\<\psi| + \frac{1}{2}|\phi\>\<\phi|)$ we have %TODO: Define f. It looks like a von Neumann entropy, but we haven't used the notation f(p) before.
\begin{align}\label{entropy}
	f(p) & = - \frac{1+p}{2} \log \frac{1+p}{2} 
	- \frac{1-p}{2} \log \frac{1-p}{2} \\
	I(\psi, \phi) &= f(p(\psi, \phi)).
\end{align}
Note that $f$ restricted to the domain $[0,1]$, the domain on which probability is defined, is invertible.
\end{proof}

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact}\footnote{%TODO: Delete this footnote?
Maybe Prop. 3.115 in \cite{moretti_spectral_2017}?} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of functions with compact support on phase space gives us finite expectation value for all functions of positions and momentum.}
